Combinatorics of the geometry of Wilson loop diagrams

### 2018-12-21 General WLD combinatorics.tex
some work on section 4

\begin{algorithm}\label{alg WLD to denom via GN}
Let $W = (\cP,n)$ be a Wilson loop diagram, and let $C(W)$ be the matrix of $W$ as defined in \eqref{C(W) dfn} (see Section~\ref{section background}).
\begin{itemize}
  \item For each $i \in [n]$, we construct a factor $r_i$ as follows:
    \begin{itemize}
      %\item Use Algorithm~\ref{alg:put GN on WLD} to obtain a bijection between the propagators and the $i-1$st Grassmann necklace element, $I_{i-1}$.  (By convention set $I_{-1}=I_n$.)  Write $I_{i-1}(p)$ for the vertex associated to propagator $p$ under this bijection.
      %\item Use Algorithm~\ref{alg:put GN on WLD} to obtain a bijection between the propagators and the $i$th Grassmann necklace element, $I_{i}$.  Write $I_{i}(p)$ for the vertex associated to propagator $p$ under this bijection.
      \item Let $S_i = \{p \in \cP \ | \ I_{i-1}(p) \neq I_i(p)\}$. (By convention, set $I_{-1} = I_n$.)
	  \item Write $\Delta_{I_i}$ for the determinant of the $k \times k$ minor of $C(W)$ with columns indexed by $I_i$.
      \item Let $r_i$ be $\Delta_{I_i}$ with all variables from rows associated to $p\not\in S_i$ set to $1$.
    \end{itemize}
  \item Define $R = \prod_{i=1}^n r_i$.
\end{itemize}
\end{algorithm}
\note{What are we doing about the sign of the determinants?  --- in the end we are making a denominator, so an object that probably makes the most sense as being defined up to sign; I would suggest having all this be up to an overall sign, but perhaps we can discuss in our next meeting.}

@@

\begin{prop}\label{prop alg gives rad}
  With notation as in Algorithm~\ref{alg WLD to denom via GN} we have the following:
  \begin{enumerate}
    \item Each $\Delta_{I_i}$ is homogeneous, as is each $r_i$.
    \item Each $\Delta_{I_i}$ splits into linear and quadratic factors.  All linear factors of  $\Delta_{I_i}$ are single variables and all irreducible quadratic factors are $2\times 2$ determinants of single variables.
    \item Quadratic factors in $r_i$ arise precisely when propagators $p$ and $q$ are supported on a common edge $(a,b)$ with $I_i(p)=a$ and $I_i(q)=b$.
    \item $r_i$ divides $\Delta_{I_i}$.
    \item The ideal generated by $R$ is the radical of the ideal generated by $\prod_{i=1}^{n}\Delta_{I_i}$.
  \end{enumerate}
\end{prop}

@@

Since there is no cancellation between terms, this means that $\Delta_{I_i}$ itself contains no other variables.  Therefore $\Delta_{I_i}$ is equal to the determinant of the matrix obtained by taking the submatrix of $C(W)$ with columns indexed by $I_i$ and setting any variables not appearing in $\Delta_{I_i}$ to $0$.  This new matrix is, up to permutations of rows and columns, a block matrix with one block for $\mathcal{P}'$ and $J$ and the other block for the complements.  Thus its determinant, and hence also $\Delta_{I_i}$, is the product of the minors for these two blocks.  By considering which variables appear, these two factors must also be $f$ and $g$, and so in particular $f$ is the minor of $C(W)$ associated to $\mathcal{P}'$ and $J$.

@@

All that remains is to prove that $\Delta_{I_i}$ has no irreducible factors of degree 3 or more.  Suppose for a contradiction that $f$ is a factor of $\Delta_{I_i}$ of degree $\geq 3$. Note that by removing the propagators which come before those contributing to $f$ and changing $i$ to be the first vertex which contributes to $f$, we obtain a different admissible diagram for which $f$ still divides $\Delta_{I_i}$ but also $i\in I_i$ and \note{would it be neater to note at the beginning that removing vertices which do not support any propagator has no effect so we assume there are no redundant vertices, and hence $i \in I_i$ for each $i$?} $i$ contributes to $f$.  Showing that this different admissible diagram gives a contradiction is sufficient, and so we may assume that $i\in I_i$ and $i$ contributes to $f$.  Finally, we can suppose that $W$ is minimal in number of propagators with the above occuring.

@@

\textbf{Case 1}: Suppose $p$ has one end on the edge $(i-1, i)$.  Thus $p$ is supported on $(i-1, i, m, m+1)$ for some $m >_i i$, and $I_{i+1}(p) = m$ by Lemma~\ref{vertex cyclic int lem}.  
%Call the propagators which appear before $p$ is closed the propagators inside $p$. \note{It might be nice to have general language for this inside/outside idea.  What do you think.}

@@

If neither $m$ nor $m+1$ appear in $I_i$ then $V(p) \cap I_i = \{i\}$,\note{check: is it obvious that $i-1 \not\in I_i$ or are we using the no coloops result?} and so the row of $p$ in the matrix of $\Delta_{I_i}$ has only one nonzero entry; hence $\Delta_{I_i}$ has a linear factor contributed by $p$ and $i$, which is a contradiction.  So we must have at least one of $m$ and $m+1$ in $I_i$.  However, all propagators in $S$ are mapped by the function $I_{i}(\cdot)$ to vertices strictly before $m$, so the matrix giving $\Delta_{I_i}$ has the form
\[
\begin{bmatrix} A & B \\ 0 & C\end{bmatrix}
\]
where $A$ is the $|S|\times |S|$ matrix indexed by the propagators in $S$ and the vertices in $I_i(S)$. No other propagators can be supported on these vertices since all other propagators are outside of $p$, and $p$ is the first propagator supported at $i$; this explains the zero block.  Therefore $\Delta_{I_i} = \det A \det C$, and both factors are nontrivial since at least one of $m$ and $m+1$ appear in $I_i$.  If we remove the propagator outside of $p$ that contributes $m$ or $m+1$, we get a smaller diagram for which $\Delta_{I_i} = \det A$. This contradicts the minimality of our choices unless $\det A$ is quadratic, which in turn contradicts our assumption that $i$ and $p$ contribute to an irreducible factor $f$ of degree at least 3.

@@

\item Suppose propagators $p$ and $q$ are supported on a common edge $(a,b)$, with $I_i(p)=a$ and $I_i(q)=b$.  Let $x_{p,a},x_{p,b},x_{q,a},x_{q,b}$ be the associated variables in $C(W)$. For any fixed bijection $\sigma$ from $\cP-\{p,q\}$ to $I_i -\{a,b\}$ for which each propagator is supported on its image under the bijection, we can extend $\sigma$ to a bijection of all propagators with $I_i$ in two ways: either $p\mapsto a$ and $q\mapsto b$ or $p\mapsto b$ and $q\mapsto a$.  The sum of the contributions of all these bijections to $\Delta_{I_i}$ is therefore the product of $x_{p,a}x_{q,b}-x_{p,b}x_{q,a}$ with the minor coming from the remaining propagators and $I_i - \{a,b\}$.  Since there is no cancellation of terms in the expansion of $\Delta_{I_i}$, if any other terms appear then they will cause a factor which is not in the form described in the previous part.  Therefore no such terms exist and $x_{p,a}x_{q,b}-x_{p,b}x_{q,a}$ is a factor of $\Delta_{I_i}$.

@@

\item Consider $p\in S_i$, and note that $\Delta_{I_i}$ is homogeneous linear in the variables of the row corresponding to $p$.  By part (2), either exactly one variable in the row corresponding to $p$ appears in $\Delta_{I_i}$ and this variable is a factor of $\Delta_{I_i}$, or exactly two variables from the row corresponding to $p$ appear in $\Delta_{I_i}$ and they appear as part of a quadratic factor.  In the first case: let the variable be $x$, then $x$ is a factor of both $r_i$ and $\Delta_{I_i}$ and is the only variable from this row in either polynomial.

@@

Next we need to show that $R$ is squarefree.  Suppose $f^2|R$.  If $f$ is a linear factor, say from associating propagator $p$ to vertex $a$, then there must be two distinct points in the Grassmann necklace algorithm where $p$ changes from not being associated to vertex $a$ to being associated to vertex $a$.  This contradicts Lemma~\ref{vertex cyclic int lem}.  Now suppose $f$ is a quadratic factor, say from associating propagators $p$ and $q$ to vertices $a$ and $b$ respectively.  \hlfix{Since $p$ comes before $q$ on edge $(a,b)$}{I can't figure out why this happens; what am I missing?} it is not possible for any $I_i$ to associate $p$ to $b$ and $q$ to $a$.  Furthermore, we know by part (4) that $p$ changes from not being associated to $a$ to being associated to $a$ if and only if $q$ changes from not being associated to $b$ to being associated to $b$.  Thus $f^2|R$ implies that twice in the Grassmann necklace $p$ must change from not being associated to vertex $a$ to being associated to vertex $a$. This is again a contradiction, and so $R$ is squarefree.

@@

\begin{thm}\label{thm denom}
Given any admissible Wilson loop diagram $W$, let $GN(W) = \{I_1, \ldots I_n\}$ be the associated Grasmann necklace. Then the denominator of the integral, $R(W)$ (see Definition~\ref{def R(W)}), is an integer multiple of the radical of $\prod_{i=1}^n \Delta_{I_i}$, where $\Delta_{I_i}$ is the determinant of the $k \times k$ minor indicated by $I_i$.
\end{thm}

\begin{proof}
In view of Proposition~\ref{prop alg gives rad} it remains to prove that the $R$ of Algorithm~\ref{alg WLD to denom via GN} is $R(W)$, the denominator of the integral $I(W)$.

To this end, first note that $R(W)$ and $R$ both have total degree $4|\mathcal{P}|$; the degree of $R(W)$ is immediate from the definition while that of $R$ follows from Lemma~\ref{vertex cyclic int lem}.  By Proposition~\ref{prop alg gives rad} every factor of $R$ is either a single variable or a quadratic factor coming from two propagators supported on a common edge and \hlfix{hence every factor of $R$ divides $R(W)$}{Is this obvious? Or does it follow from a previous result?}.  Finally, since $R$ is squarefree and monic, this implies that $R(W)$ is an integer multiple of $R$.
\end{proof}

>>>>

\begin{algorithm}\label{alg WLD to denom via GN}
Let $W = (\cP,n)$ be a Wilson loop diagram, and let $C(W)$ be the matrix of $W$ as defined in \eqref{C(W) dfn} (see Section~\ref{section background}).
\begin{itemize}
  \item For each $i \in [n]$, we construct a factor $r_i$ as follows:
    \begin{itemize}
      %\item Use Algorithm~\ref{alg:put GN on WLD} to obtain a bijection between the propagators and the $i-1$st Grassmann necklace element, $I_{i-1}$.  (By convention set $I_{-1}=I_n$.)  Write $I_{i-1}(p)$ for the vertex associated to propagator $p$ under this bijection.
      %\item Use Algorithm~\ref{alg:put GN on WLD} to obtain a bijection between the propagators and the $i$th Grassmann necklace element, $I_{i}$.  Write $I_{i}(p)$ for the vertex associated to propagator $p$ under this bijection.
      \item Let $S_i = \{p \in \cP \ | \ I_{i-1}(p) \neq I_i(p)\}$. (By convention, set $I_{-1} = I_n$.)
	  \item Write $\Delta_{I_i}$ for the determinant of the $k \times k$ minor of $C(W)$ with columns indexed by $I_i$.
      \item Let $r_i$ be $\Delta_{I_i}$ with all variables from rows associated to $p\not\in S_i$ set to $1$.
    \end{itemize}
  \item Define $R = \prod_{i=1}^n r_i$.
\end{itemize}
\end{algorithm}

As we take each $\Delta_{I_i}$ to simply be the determinant of a particular submatrix of $C(W)$, the overall sign of each $\Delta_{I_i}$ is well-defined.  However, we will not be interested in these signs as they only contribute an overall sign to $R$ and our main theorem in this section (Theorem~\ref{thm denom}) will simply be to show that $R$ is equal up to a $\mathbb{Q}$-multiple to the denominator of the Wilson loop diagram (see Definition~\ref{def R(W)}).

@@

\begin{prop}\label{prop alg gives rad}
  With notation as in Algorithm~\ref{alg WLD to denom via GN} and working over the field $\mathbb{Q}$ we have the following:
  \begin{enumerate}
    \item Each $\Delta_{I_i}$ is homogeneous, as is each $r_i$.
    \item Each $\Delta_{I_i}$ splits into linear and quadratic factors.  All linear factors of  $\Delta_{I_i}$ are single variables and all irreducible quadratic factors are $2\times 2$ determinants of single variables.
    \item Quadratic factors in $\Delta_{I_i}$ arise precisely when propagators $p$ and $q$ are supported on a common edge $(a,b)$ with $I_i(p)=a$ and $I_i(q)=b$.
    \item $r_i$ divides $\Delta_{I_i}$.
    \item The ideal generated by $R$ is the radical of the ideal generated by $\prod_{i=1}^{n}\Delta_{I_i}$.
  \end{enumerate}
\end{prop}

@@

Since there is no cancellation between terms, this means that the full expansion over permutations of $\Delta_{I_i}$ contains no other nonzero terms and hence no other variables.  Therefore $\Delta_{I_i}$ is equal to the determinant of the matrix obtained by taking the submatrix of $C(W)$ with columns indexed by $I_i$ and setting any variables not appearing in $\Delta_{I_i}$ to $0$.  This new matrix is, up to permutations of rows and columns, a block matrix with one block for $\mathcal{P}'$ and $J$ and the other block for the complements.  Thus its determinant, and hence also $\Delta_{I_i}$, is the product of the minors for these two blocks.  By considering which variables appear, these two factors must also be $f$ and $g$, and so in particular $f$ is the minor of $C(W)$ associated to $\mathcal{P}'$ and $J$.  This proves the claim.

@@

All that remains is to prove that $\Delta_{I_i}$ has no irreducible factors of degree 3 or more.  Suppose for a contradiction that $f$ is a factor of $\Delta_{I_i}$ of degree $\geq 3$. Note that by removing the propagators which come before those contributing to $f$ and changing $i$ to be the first vertex which contributes to $f$, we obtain a different admissible diagram for which $f$ still divides $\Delta_{I_i}$ but also $i\in I_i$ and \note{would it be neater to note at the beginning that removing vertices which do not support any propagator has no effect so we assume there are no redundant vertices, and hence $i \in I_i$ for each $i$?  No, the point is that we can make it so that $i$ is the first index where a propagator of $f$ contributes, not just that it is in $I_i$.} $i$ contributes to $f$.  Showing that this different admissible diagram gives a contradiction is sufficient, and so we may assume that $i\in I_i$ and $i$ contributes to $f$.  Finally, we can suppose that $W$ is minimal in number of propagators with the above occuring.

@@

\textbf{Case 1}: Suppose $p$ has one end on the edge $(i-1, i)$.  Thus $p$ is supported on $(i-1, i, m, m+1)$ for some $m >_i i$, and $I_{i+1}(p) = m$ by Lemma~\ref{vertex cyclic int lem}.

@@

If neither $m$ nor $m+1$ appear in $I_i$ then by Corollary~\ref{no coloops} $V(p) \cap I_i = \{i\}$, and so the row of $p$ in the matrix of $\Delta_{I_i}$ has only one nonzero entry; hence $\Delta_{I_i}$ has a linear factor contributed by $p$ and $i$, which is a contradiction.  So we must have at least one of $m$ and $m+1$ in $I_i$.  However, all propagators in $S$ are mapped by the function $I_{i}(\cdot)$ to vertices strictly before $m$, so the matrix giving $\Delta_{I_i}$ has the form
\[
\begin{bmatrix} A & B \\ 0 & C\end{bmatrix}
\]
where $A$ is the $|S|\times |S|$ matrix indexed by the propagators in $S$ and the vertices in $I_i(S)$. No other propagators can be supported on these vertices since all other propagators are outside of $p$, and $p$ is the first propagator supported at $i$; this explains the zero block.  Therefore $\Delta_{I_i} = \det A \det C$, and both factors are nontrivial since at least one of $m$ and $m+1$ appear in $I_i$.  If we remove the propagator outside of $p$ that contributes $m$ or $m+1$, we get a smaller diagram for which $\Delta_{I_i} = \det A$. This contradicts the minimality of our choices unless $\det A$ is quadratic, which in turn contradicts our assumption that $i$ and $p$ contribute to an irreducible factor $f$ of degree at least 3.

@@

\item Suppose propagators $p$ and $q$ are supported on a common edge $(a,b)$, with $I_i(p)=a$ and $I_i(q)=b$.  Let $x_{p,a},x_{p,b},x_{q,a},x_{q,b}$ be the associated variables in $C(W)$. For any fixed bijection $\sigma$ from $\cP-\{p,q\}$ to $I_i -\{a,b\}$ for which each propagator is supported on its image under the bijection, we can extend $\sigma$ to a bijection of all propagators with $I_i$ in two ways: either $p\mapsto a$ and $q\mapsto b$ or $p\mapsto b$ and $q\mapsto a$.  The sum of the contributions of all these bijections to $\Delta_{I_i}$ is therefore the product of $x_{p,a}x_{q,b}-x_{p,b}x_{q,a}$ with the minor coming from $\cP-\{p, q\}$ and $I_i - \{a,b\}$.  Since there is no cancellation of terms in the expansion of $\Delta_{I_i}$, if any other terms appear then they will cause a factor which is not in the form described in the previous part.  Therefore no such terms exist and $x_{p,a}x_{q,b}-x_{p,b}x_{q,a}$ is a factor of $\Delta_{I_i}$.

@@

\item Consider $p\in S_i$, and note that $\Delta_{I_i}$ is homogeneous linear in the variables of the row corresponding to $p$.  By part (2), either exactly one variable in the row corresponding to $p$ appears in $\Delta_{I_i}$ and this variable is a factor of $\Delta_{I_i}$, or exactly two variables from the row corresponding to $p$ appear in $\Delta_{I_i}$ and they appear as part of a quadratic factor.  In the first case let the variable be $x$, then $x$ is a factor of both $r_i$ and $\Delta_{I_i}$ and is the only variable from this row in either polynomial.

@@

Next we need to show that $R$ is squarefree.  Suppose $f^2|R$.  If $f$ is a linear factor, say from associating propagator $p$ to vertex $a$, then there must be two distinct points in the Grassmann necklace algorithm where $p$ changes from not being associated to vertex $a$ to being associated to vertex $a$.  This contradicts Lemma~\ref{vertex cyclic int lem}.  Now suppose $f$ is a quadratic factor, say from propagators $p$ and $q$ supported on the edge $(a, b)$ with $p$ before $q$ on the edge.  In this case it is not possible for any $I_i$ to associate $p$ to $b$ and $q$ to $a$.  Furthermore, we know by part (4) that $p$ changes from not being associated to $a$ to being associated to $a$ if and only if $q$ changes from not being associated to $b$ to being associated to $b$.  Thus $f^2|R$ implies that twice in the Grassmann necklace $p$ must change from not being associated to vertex $a$ to being associated to vertex $a$. This is again a contradiction, and so $R$ is squarefree.

@@

\begin{thm}\label{thm denom}
Given any admissible Wilson loop diagram $W$, let $\{I_1, \ldots I_n\}$ be the associated Grasmann necklace. Then the denominator of the integral, $R(W)$ (see Definition~\ref{def R(W)}), is a $\mathbb{Q}$-multiple of the radical of $\prod_{i=1}^n \Delta_{I_i}$, where $\Delta_{I_i}$ is the determinant of the $k \times k$ minor indicated by $I_i$.  Equivalently, $R(W)$ is a $\mathbb{Q}$-multiple of the $R$ of Algorithm~\ref{alg WLD to denom via GN}.
\end{thm}

\begin{proof}
The equivalence of the final statement of the theorem is due to Proposition~\ref{prop alg gives rad}.  It remains to prove that $R(W)$ is an integer multiple of the $R$ of Algorithm~\ref{alg WLD to denom via GN}.

To this end, first note that $R(W)$ and $R$ both have total degree $4|\mathcal{P}|$; the degree of $R(W)$ is immediate from the definition while that of $R$ follows from Lemma~\ref{vertex cyclic int lem}.  By Proposition~\ref{prop alg gives rad} every factor of $R$ is either a single variable or a quadratic factor coming from two propagators supported on a common edge.  The factors of each $R_e$ making up $R(W)$ in the notation of Definition~\ref{def R(W)} are all of this form and hence every factor of $R$ divides $R(W)$.
Finally, since $R$ is squarefree, this implies that $R(W)$ is a $\mathbb{Q}$-multiple of $R$.
\end{proof}

\note{Is there any point in working out the $\mathbb{Q}$-coefficient, or otherwise discussing the field?}
